The Bolocam Galactic Plane Survey: Survey Description and Data Reduction

### 2009-04-13 methods/ms.tex
Added Neal's introduction, John's cygnus comparison, and comment on distortion

\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{{\tt jaguirre@sas.upenn.edu}}
\newcommand{\lsim}{{_{<}\atop^{\sim}}}
\newcommand{\gsim}{{_{>}\atop^{\sim}}}
\newcommand{\etal}{{et al.\/}}
\newcommand{\ie}{{\em ie.\/}}
\newcommand{\cmq}{cm{$^{-3}$}}
\newcommand{\per}{$^{\rm{-1}}$}
\newcommand{\tc}{{$\theta^1$~Orionis~C}}
%\newcommand{\msol}{M{$_{\odot}$}}
\def\msol{\ifmmode {\>M_\odot}\else {$M_\odot$}\fi}
\newcommand{\lsol}{L{$_{\odot}$}}
\newcommand{\kms}{km~s{$^{-1}$}}
\newcommand{\hii}{H~{\sc ii}}
\newcommand{\Hii}{H~{\sc ii}}
\newcommand{\Ha}{\mbox{H$\alpha$}}
\newcommand{\sii}{S~{\sc ii}}
\newcommand{\Feii}{Fe~{\sc ii}}
\newcommand{\oi}{O~{\sc i}}
\newcommand{\nii}{N~{\sc ii}}
\newcommand{\oiii}{O~{\sc iii}}
\newcommand{\mgii}{Mg~{\sc ii}}
\newcommand{\tco}{{$^{13}$CO}}
\newcommand{\CO}{{$^{12}$CO}}
\newcommand{\Tco}{{$^{12}$CO}}
\newcommand{\co}{C{$^{18}$O}}
\newcommand{\Lsol}{L$_{\odot}$}
\newcommand{\Msol}{M$_{\odot}$}
%\newcommand{\C18o}{C$^{18}$O($1\rightarrow 0$)}
\newcommand{\Check}{{\bf ???}}
\newcommand{\mum}{\ensuremath{\mu \mathrm{m}}}
\newcommand{\flux}{flux density}
\newcommand{\solar}{\ensuremath{\odot}}

@@

Millimeter-wavelength continuum surveys of the Galactic plane provide
the best way to identify high-column density clumps and clumps where
planets, stars, and star clusters form.  Such data can locate and lead
to the measurement of the physical properties of clumps un-biased by
selection effects such as the presence of embedded stars, star
clusters, infrared sources, masers, or radio continuum emission.
Galaxy-wide surveys are essential for measuring the impacts of the
environment on clump properties and star formation activity.  Do clump
properties vary with location in or out of spiral arms?  Do clump
properties vary with Galactocentric distance?  Do they depend on the
level of nearby star formation activity?  Galactic plane surveys of
mm-wavelength dust continuum emission provide the most efficient tool
for the identification of potential or active sites of star formation,
including the rare objects where the most massive stars and cluster
are, or soon will be, forming.  Dust continuum surveys of the Galaxy
provide essential ``ground truth'' required for the analysis of distant
galaxies where where individual clouds and clumps are not resolved and
only galaxy-wide average quantities can be measured.

Dust continuum emission near a wavelength of 1 mm is the best tracer
of the material most directly associated with the formation of
planetary systems, stars, and star clusters \citep{johnstone06}.
Galactic dust continuum radiation at this wavelength is optically-thin
almost everywhere, and at typical dust temperatures of order 10 K or
more, is observed on the Rayleigh-Jeans tail of the Planck function.
Thus, the determination of column densities is relatively straight
forward if the dust temperature is known.

Submillimeter observations of Galactic sources have been conducted by
a variety of different groups focusing on small regions since
bolometer technology was developed.  The Submillimeter Common-User
Bolometer Array (SCUBA) on the 15 m James Clerk Maxwell Telescope
mapped many significant regions in the galactic plane over its 8 year
lifetime \citep{holland99,difrancesco08}, but never completed a
contiguous survey of the galactic plane.  Bolocam and Mambo have
previously been used to map large star forming regions at 1.1
\citep{enoch06} and 1.2 mm \citep{motte07}, but again these were
targeted observations.  A recent large area survey with the LABOCA
instrument at 870 \mum\ has recently been reported \citep{schuller09}.

Surveys of the galaxy in CO line emission \citep{dame01,FCRAO} have
revealed much about galactic structure, but do not provide a complete
picture of how star formation follows that structure.  Millimeter
continuum more directly traces the cold dust associated with star
formation.  Continuum data allows for accurate determinations of the
mass of large, dense dust structures when combined with kinematic
distances \citep[e.g.][]{RotationCurve}.

A crucial step in the observational study of star and cluster
formation is identification and characterization of the cloud clumps
which will soon form or are actively forming stars.  Massive stars and
clusters form from high-density cloud clumps with very large column
densities and extinctions $A_V > $ 100.  Thus, such cloud clumps are
best investigated at millimeter and sub-millimeter wavelengths.
However, the interpretation of various gas tracers that produce
emission lines in this portion of the spectrum is very difficult.
Although in principle, spectral lines provide excellent diagnostics of
line-of-sight motions, temperatures, and densities in a cloud,
variations in tracer abundances caused by depletions and complex
chemical processing, uncertainties in excitation conditions, and the
impacts of radiation fields and shocks make the derivation of column
densities, masses, and other physical properties of cloud clumps very
difficult.  The continuum emission from warm dust provides a more
reliable tracer of the column densities and clump masses.

The advent of focal plane arrays containing hundreds of individual
bolometers sensitive to millimeter and sub-millimeter (sub-mm)
radiation has enabled large-scale surveys of mm-wavelength continuum
emission from Galactic interstellar dust.  The great advantage of
continuum observations at 1.1mm is the low optical depth of the dust,
permitting an estimate of the mass of the emitting region.  Further,
the mass column density is independent of the gas volume density.
Thus it is thus possible with a wide area, relatively shallow 1.1~mm
survey to provide an unmatched, uniform inventory of massive
star-forming and starless condensations.

This survey presents the results of one of the first wide-area,
untargeted surveys of the Galactic Plane in the millimeter continuum.
We have used the 144 element Bolocam focal plane array, mounted at the
Cassegrain focus of the 10 meter diameter sub-mm telescope at the
Caltech Sub-millimeter Observatory (CSO), to survey over \bgpsarea\
square degrees of the northern Galactic plane at a wavelength of 1.1
mm in the dust continuum.  Figure \ref{fig:Coverage} shows the
coverage of the \bgps.  Maps of the entire BGPS are presented in
Figure \ref{fig:BGPSMontage}.  We have detected \ncores\ clumps in the
surveyed fields, providing an unbiased sample of clumps suitable for
multi-wavelength and high-resolution studies with existing telescopes
and future facilities.  At the distances that we are observing these
objects, we are likely observing aggregates of star forming cores
found in a larger molecular cloud environment, which we will refer to
as clumps \citep{williams00}.  The clump catalog is described in a
companion paper \citep{rosolowsky09}.

@@

The Bolocam and M07 data sets have flux calibration consistent within
about 20\%.  However, direct comparison with the fluxes listed in M07
is difficult due to the factor of three difference in the size of the
beams used.  For example, the the brightest source in the Cyg-X region
is DR 21.  While M07 quote a peak flux density per beam of
4.2 Jy/beam for the brightest feature in the DR 21 complex (their peak
N46) and 17.9 Jy/beam in a 0.19 x 0.14 pc box, BGPS measures a peak
flux of 15.5 Jy/beam (the beam being roughly 31\arcsec\ in diameter).
Fluxes are best compared using sources compact as observed with the
beam of the IRAM 30 meter telescope.

We have convolved the fits files\footnote{{\tt
ftp://cdsarc.u-strasbg.fr/pub/cats/J/A+A/476/1243/fits/}} with a
Gaussian kernel that generates an approximately 35\arcsec\ beam,
matched to our images (mapped without application of the distortion
correction).  The results are shown in Figure
\ref{fig:CygnusComparison}.

\Table
{lll}
{Flux density comparison with \citet{motte07}.}
{
Source  &   BGPS           &  Motte \\
        &   peak mJy/beam  &  mJy in box (see Motte+ Table 1)
}
{tab:CygnusComparison}
{
 N11    &   350 - 550      &  340 \\
 N12    &   1660-2200      &  1630 \\
 N16    &   613            &  550 \\
 N18    &   618            &  430 \\
 N26    &   314            &  340 \\
 N46    &   15500          &  17930  (DR21 S)
}

@@

>>>>

\newcommand{\ammonia}{\mbox{{\rm NH}$_3$}}
\newcommand{\form}{H$_2$CO}
\newcommand{\td}{\mbox{$T_D$}}
\newcommand{\jj}[2]{\mbox{$J = #1\rightarrow#2$}}
\newcommand{\vdag}{(v)^\dagger}
\newcommand{\myemail}{{\tt jaguirre@sas.upenn.edu}}
\newcommand{\lsim}{{_{<}\atop^{\sim}}}
\newcommand{\gsim}{{_{>}\atop^{\sim}}}
\newcommand{\etal}{{et al.\/}}
\newcommand{\ie}{{\em ie.\/}}
\newcommand{\cmq}{cm{$^{-3}$}}
\newcommand{\per}{$^{\rm{-1}}$}
\newcommand{\tc}{{$\theta^1$~Orionis~C}}
%\newcommand{\msol}{M{$_{\odot}$}}
\def\msol{\ifmmode {\>M_\odot}\else {$M_\odot$}\fi}
\newcommand{\lsol}{L{$_{\odot}$}}
\newcommand{\kms}{km~s{$^{-1}$}}
\newcommand{\hii}{H~{\sc ii}}
\newcommand{\Hii}{H~{\sc ii}}
\newcommand{\Ha}{\mbox{H$\alpha$}}
\newcommand{\sii}{S~{\sc ii}}
\newcommand{\Feii}{Fe~{\sc ii}}
\newcommand{\oi}{O~{\sc i}}
\newcommand{\nii}{N~{\sc ii}}
\newcommand{\oiii}{O~{\sc iii}}
\newcommand{\mgii}{Mg~{\sc ii}}
\newcommand{\tco}{{$^{13}$CO}}
\newcommand{\CO}{{$^{12}$CO}}
\newcommand{\Tco}{{$^{12}$CO}}
\newcommand{\co}{C{$^{18}$O}}
\newcommand{\Lsol}{L$_{\odot}$}
\newcommand{\Msol}{M$_{\odot}$}
%\newcommand{\C18o}{C$^{18}$O($1\rightarrow 0$)}
\newcommand{\Check}{{\bf ???}}
\newcommand{\mum}{\ensuremath{\mu \mathrm{m}}}
\newcommand{\flux}{flux density}
\newcommand{\solar}{\ensuremath{\odot}}

@@

Millimeter-wavelength continuum surveys of the Galactic plane provide
the most efficient way to find molecular clumps that are the likely sites of
formation for massive stars and star clusters. The development of
detector arrays have made blind surveys of large areas possible.
Such surveys can remove the selection effects used in previous surveyes,
such as the presence of embedded stars or star clusters, infrared sources,
masers, or radio continuum emission. In particular, they can locate
molecular clumps before stars form, providing vital information on
the initial conditions.  The surveys can also provide valuable constraints
on the physical properties of the clumps, especially masses and mean densities.

Galaxy-wide surveys are essential for measuring the impacts of the
environment on clump properties and star formation activity.
Do clump properties vary with Galactocentric distance, or
with location with respect to spiral arms?  Do they depend on the
level of nearby star formation activity?
Answering these questions in our Galaxy will provide the
essential ``ground truth'' required for the analysis of distant
galaxies where where individual clouds and clumps are not resolved and
only galaxy-wide average quantities can be measured.

Surveys of the galaxy in CO line emission \citep{dame01,FCRAO} have
identified the locations of large molecular clouds and traced
the transition from atomic to molecular gas as distance from the
center of the galaxy decreases. Together with kinematic distances
\citep[e.g.][]{RotationCurve} and, more recently, accurate distances
from VLBI (add ref to Reid et al. work), these surveys have begun to
reveal the spiral structure of the Milky Way. As this picture comes into
sharper focus, we can begin to address the old question of the relationship
between spiral density waves and star formation.

To pursue this question, we need tracers of higher density material.
Studies of nearby clouds have demonstrated that star formation is far
from uniform over molecular clouds, but is concentrated in unusually dense
regions called clumps or cores, which fill a small fraction of the area
of most clouds [refs. E. Lada et al. Evans et al., Enoch et al.]
Following \citet{williams00} and McKee-Ostriker (ARAA), we
use the term core to refer to a very dense region destined to form an
individual star or small multiple star systems and the term clump to
refer to a
region likely to form a group or cluster of stars. While clumps forming
massive clusters may be denser than cores forming low mass stars
[ref Mueller?],
in general, clumps are larger and less dense. Clumps can be traced
with line emission from molecules such as \ammonia, \form, CS, HCN, etc.
Spectral lines also provide excellent diagnostics of
line-of-sight motions, temperatures, and densities.
However, the interpretation of these gas tracers is complicated by
variations in tracer abundances caused by depletions and complex
chemical processing, uncertainties in optical depths and excitation
conditions, and the impacts of radiation fields and shocks. These
complications make the derivation of column densities, masses, and
other physical properties of star-forming clumps very difficult.
A further limitation of spectral line studies is that the tracers of
denser gas tend to be weak and multi-element receiver arrays have
only a modest number of elements, making blind surveys very time-consuming.

In contrast, focal-plane arrays containing hundreds of individual
bolometers sensitive to millimeter and sub-millimeter (sub-mm)
radiation are now available. This advance in technology enables
blind surveys of the Galactic plane, providing a uniform inventory of massive
star-forming and starless clumps.  Catalogs resulting
from such surveys will provide the data base for subsequent observations
in tracers such as hard-to-excite molecular lines and higher resolution
studies with instruments like CARMA and ALMA.
A further advantage of continuum observations of dust at long wavelengths
is the low optical depth of the dust \citep{johnstone06},
permitting a straight-forward estimate of the mass of the emitting region.

Submillimeter observations of Galactic sources have been conducted by
a variety of different groups focusing on small regions since
bolometer technology was developed.  The Submillimeter Common-User
Bolometer Array (SCUBA) on the 15-m James Clerk Maxwell Telescope
mapped many significant regions in the galactic plane over its 8 year
lifetime \citep{holland99,difrancesco08}, but never completed a
contiguous survey of the galactic plane.  The Bolocam instrument has been
used on the 10.4-m Caltech Submiliimeter Observatory (CSO) to map nearby large
molecular clouds at 1.1 mm and identify hundreds of cores \citep{enoch06}.
Others have used MAMBO on the 30-m IRAM telescope at 1.2 mm to map
more distant regions of more massive star formation \citep{motte07}.
None of these surveys covered a significant fraction of the Galactic plane.

Two large surveys of the Galactic plane are in progress.
The ATLASGAL survey, using the LABOCA instrument at 870 \micron\
on the APEX telescope in Chile, will cover the southern Galactic plane
and extend somewhat into the north. Initial results from this survey
have been reported by \citet{schuller09}. [say more]

This paper presents the results of the Bolocam Galactic Plane Survey
(BGPS).  (Give longitude and latitude coverage.)
It is complementary to the survey by \citet{schuller09} in that
it targets the part of the plane visible from the northern
hemisphere and uses a somewhat longer wavelength, allowing
comparison of properties in the region of overlap with the survey of
\citet{schuller09}.
At 870 \micron, $h\nu/k = 16.5$ K, compared to 13.1 K at 1.1 mm.
Because starless clumps may be quite cold ($\td \approx 10$ K),
the uncertainty in mass caused by uncertain temperature is less at 1.1 mm.
For example, the mass estimate if the actual $\td = 20$ K, rather
than 10 K, would decrease by a factor of 3.3 at 870 \micron\
versus a factor of 2.9 at 1.1 mm. In addition, the Bolocam bandpass
excludes the CO \jj{21} line, which can be a significant contaminant
on top of the dust continuum emission.

We have used the 144-element Bolocam focal plane array, mounted at the
Cassegrain focus of the CSO, to survey over \bgpsarea\
square degrees of the northern Galactic plane at a wavelength of 1.1
mm in the dust continuum.  Figure \ref{fig:Coverage} shows the
coverage of the \bgps.  Maps of the entire BGPS are presented in
Figure \ref{fig:BGPSMontage}.  We have detected \ncores\ clumps in the
surveyed fields, providing an unbiased sample of clumps suitable for
multi-wavelength and high-resolution studies with existing telescopes
and future facilities.  At the distances that we are observing these
objects, we are likely observing aggregates of star forming cores
found in a larger molecular cloud environment, which we will refer to
as clumps \citep{williams00}.  The clump catalog is described in a
companion paper \citep{rosolowsky09}.

@@

The Bolocam and M07 data sets have flux calibration consistent within
about 20\%.  However, direct comparison with the fluxes listed in M07
is difficult due to the factor of three difference in the size of the
beams used.  For example, the the brightest source in the Cyg-X region
is DR 21.  While M07 quote a peak flux density per beam of
4.2 Jy/beam for the brightest feature in the DR 21 complex (their peak
N46) and 17.9 Jy/beam in a 0.19 x 0.14 pc box, BGPS measures a peak
flux of 17.5 Jy/beam (the beam being roughly 35\arcsec\ in diameter).
Fluxes are best compared using sources compact as observed with the
beam of the IRAM 30 meter telescope.

We have convolved the fits files\footnote{{\tt
ftp://cdsarc.u-strasbg.fr/pub/cats/J/A+A/476/1243/fits/}} with a
Gaussian kernel that generates an approximately 35\arcsec\ beam,
matched to our images.  The results are shown in Figure
\ref{fig:CygnusComparison}.

\Table
{llllll}
{Note:  The \citet{motte07} integration box is given in terms of pc. 
The assumed distance must be about 1.7 kpc \citep{schneider2006}  }
{
%Source  &   BGPS           &  Motte \\
%        &   peak mJy/beam  &  mJy in box (see Motte+ Table 1)
Motte Source Number &  Motte Table 1 integrated flux &  Motte scaled Jy/bm  & BGPS Jy/Bm &    Comments \\
}
{tab:CygnusComparison}
{
3  &    1.59 [0.11x0.09] & 2.61      &  2.30  &  DR 17                               \\
11 &    0.34 [0.11x0.06] & 0.48      &  0.50  &  West SW of DR21                     \\
12 &    1.63 [0.11x0.09] & 2.60      &  2.20  &    "                                 \\
16 &    0.55 [0.12x0.12] & 0.82      &  0.72  &    "                                 \\
26 &    0.34 [0.20x0.10] & 0.48      &  0.44  &    "                                 \\
29 &    0.54 [0.25x0.13] & 1.10      &  0.62  &    "                                 \\
30 &   10.54 [0.15x0.10] &13.5       &  9.80  &  W75(N)                              \\
44 &    8.43 [0.14x0.11] &14.5       & 12.7   &  DR21 (OH)                           \\
46 &   17.90 [0.08x0.03] &18.4       & 15.8   &  DR21 f-f, maser (H2O)               \\
48 &    3.72 [0.17x0.11] & 9.4       &  7.8   &  DR21 (OH)-S                         \\
51 &    2.37 [0.24x0.03] & 5.5       &  4.1   &  DR21 FIR2                           \\
53 &    1.60 [0.14x0.09] & 4.3       &  2.7   &  North of DR21  FIR3                 \\
56 &    0.60 [0.12x0.08] & 0.81      &  0.75  &  SE of DR21 47                       \\
58 &    2.73 [0.23x0.19] & 2.40      &  1.64  &  IRAS 20375+4109                     \\
59 &    0.37 [0.12x0.05] & 1.3       &  0.93  &  East of IRAS 20375+4109             \\
60 &    0.56 [0.16x0.13] & 1.00      &  0.78  &  East of IRAS 20375+4109             \\
61 &    0.23 [0.12x0.09] & 0.60      &  0.49  &  MSX 81.5168+0.01926                 \\
62 &    0.50 [0.18x0.12] & 1.23      &  0.80  &  between 63 \& IRAS 20375+4109       \\
63 &    1.09 [0.08x0.05] & 1.60      &  1.50  &  DR22                                \\
65 &    0.78 [0.16x0.12] & 1.15      &  1.10  &                                      \\
66 &    0.25 [0.15x0.06] & 0.63      &  0.76  &                                      \\
68 &    0.84 [0.10x0.10] & 1.01      &  0.94  &                                      \\
69 &    1.28 [0.47x0.18] & 1.12      &  1.20  &                                      \\
72 &    0.54 [0.28x0.13] & 0.58      &  0.47  &                                      \\
}

@@

In our first release, these positions are made using the nominal positions
without correction for optical distortion.  This results in a slightly blurred
beam (35.6\farcs instead of 31.2\farcs).  Because this distortion is
asymmetric, it can also result in a small pointing offset at the field edges
where only bolometers with large offsets record data.  However, this effect is
strongly mitigated by the low number of hits at field edges - this data is
generally flagged out.  The residual offset is within the overall pointing
model error, but caution should be used when measuring locations at the field
edges.
